Double Standards in Judging Collective Action

### 2022-08-16 Documents/manuscript.Rmd
first complete draft

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
keywords: protest, collective action, double standards, social identity theory, system justification theory, Black Lives Matter
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In four studies (total *N* = 2,246), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In two preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own (system-challenging) ideological position---but participants showed no ingroup bias when judging collective action by ingroup and outgroup members. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
bibliography: references.bib
csl: apa.csl
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: yes
    template: preprint-template.tex

@@

<!-- I should cite Colin Leach's work. -->

Protest movements often spark contention about what actions are acceptable means of protest. Reactions to Black American athletes kneeling during the national anthem to protest racist police violence are a case in point. Kneeling during the national anthem is not violent, disruptive, or illegal. And yet, only 29\% of White Americans (66\% of Black Americans) and 11\% of Republicans (59\% of Democrats) considered it appropriate for Black athletes to protest in this way [@yougov_nfl_2017]. Indeed, Black athletes who had participated in the protests faced a backlash in public opinion [@lacina_nfl_2020]. This and other examples suggest that *who* the protesters are and *what* they are protesting influences how acceptable observers judge their actions to be.

@@

The second purpose of our research is to test for double standards in judging collective action---that is, whether people judge the same protest actions as more acceptable if the protesters' are ingroup members or the protesters' cause aligns with the ingroup's interests (Hypothesis 1a, 1b) or if the protesters' cause aligns with their own system-justifying or system-challenging ideological orientation (Hypotheses 2a, 2b). In Experiment 1, we use the instrument developed in Studies 1 and 2 to test for double standards in how acceptable participants consider collective action for workers' rights in the United Kingdom. By varying both the participants' and the protesters' group memberships (working/middle class), Experiment 1 compares the relative evidence for identity-based (Hypotheses 1a, 1b) and ideology-based double standards (Hypothesis 2a). Extending Experiment 1, Experiment 2 considers both system-challenging and system-defending collective action in the context of the 2020 Black Lives Matter protests in the United States. By varying the protesters' cause (for/against defunding the police) as well as the participants' and the protesters' group memberships (Black/White), Experiment 2 compares the relative evidence identity-based (Hypotheses 1a, 1b) and ideology-based double standards (Hypothesis 2a, 2b). Together, the preregistered experiments provide a comprehensive test of the hypothesized double standards in judging collective action.

@@

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that, in line with Hypothesis 1a and 1b, participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights than if they were outgroup members protesting for the outgroup's rights. By comparing reactions to protesters with working-class or professional jobs, we tested how observers judge collective action by lower-status (i.e. working-class) and higher-status (professional/middle-class) group members. Assuming that collective action by lower-status group members is more threatening to the unequal status than collective action by higher-status group members, we tested the preregistered hypothesis that, in line with Hypothesis 2a, participants would judge the same protest actions as more acceptable if the protesters were from a higher-status group than if they were from a lower-status group. In non-preregistered analyses, we further explored whether, in line with Hypothesis 2b, participants would judge the same protest actions as more or less acceptable depending on their own ideological positions. In this way, Experiment 1 provided a first test of the hypothesized identity- and ideology-based double standards in judging collective action.

@@

We preregistered the sample size as well as all hypotheses, inclusion/exclusion criteria, measures, and manipulations.^[[https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b](https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b)] We made all materials, data, and analysis scripts available online.^[[https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)] We followed sample size recommendations for item response theory models [@demars_irt_2010, p. 36], planning to recruit 500 participants.

@@

We recruited 500 participants from the remaining 687 participants, 250 from the lower-status group and 250 from the higher-status group. Participants were randomly assigned to read a vignette about a government bill affecting either people in working-class jobs (lower-status protesters) or people in professional jobs (higher-status protesters). Participants in both conditions were instructed to carefully read the vignette and to try to imagine what it would be like if this situation was real. Participants in the lower-status protestprs condition read the following introduction:

@@

Figure \ref{fig:f3} (A) shows estimates for each combination of the protesters' and the participants' group membership. Overall, we found that participants' responses depended on both the protesters' and the participants' group status---but not in the directions predicted by our hypotheses. Contradicting Hypothesis 1, participants did not consider protest actions performed by their ingroup to be more acceptable, on average, than the same actions performed by the relevant outgroup (Cohen's $d = 0.00, [-0.20, 0.20]$). Instead, we found that participants from higher-status backgrounds considered protest actions performed by both lower-status ($d = 0.34, [0.07, 0.59$]) and higher-status ($d = 0.33, [0.03, 0.63]$) protesters to be, on average, more acceptable than participants from lower-status backgrounds. Contradicting Hypothesis 2, participants considered protest actions performed by higher-status protesters to be less acceptable, on average, than the same actions performed by lower-status protesters ($d = -0.20, [-0.40, -0.00]$).

@@

\ldots

@@

In Experiment 2, we investigated potential double standards in judging collective action for or against defunding the police in the United States. That is, we tested whether Black and White Americans applied different standards when judging collective action taken by either Black or White protesters to protest either for or against police divestment as a possible solution to racist police violence.

- Explain why defunding the police (public support for BLM, so needed to be more controversial)

<!-- In Experiment 2, we tested five hypotheses. We tested two predictions derived from social identity theory. First, we hypothesize that observers are more likely to judge an action as an acceptable means of protest when the protesters are ingroup members than when they are outgroup members (Hypothesis 1a). Second, we hypothesize that observers are more likely to judge an action as an acceptable means of protest when the goal of the protests would benefit the observers' ingroup (Hypothesis 1b). We will test two predictions derived from system justification theory. First, we hypothesize that observers are more likely to judge an action as an acceptable means of protest when the goal of the protest is to support rather than challenge the system (Hypothesis 2a). Second, we hypothesize that observers who endorse system-justifying beliefs are more likely to consider system-supporting actions acceptable means of protest-and less likely to consider system-challenging actions acceptable means of protest-than observers who do not endorse system-justifying beliefs (Hypothesis 2b). In addition, we will test an alternative hypothesis. We hypothesize that observers are more likely to consider actions as acceptable means of protest when they support the cause of the protesters than when they oppose their cause (Hypothesis 3). We note that Hypotheses 1a, 1b, and 2a make competing predictions about differences between conditions while Hypotheses 2b and 3 need not contradict each other or any of the other hypotheses. -->

@@

We preregistered the sample size as well as all hypotheses, inclusion/exclusion criteria, measures, and manipulations.^[[https://osf.io/skxjt/?view_only=d8ce44a700884b5ab36b64ef08f833a1](https://osf.io/skxjt/?view_only=d8ce44a700884b5ab36b64ef08f833a1)] We made all materials, data, and analysis scripts available online.^[[https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)] As reported in the preregistration, we ran simulations, using data from Experiment 1, to determine that a sample size of $N = 1,600$ ($n = 200$ per condition) was sufficient to detect even small differences between conditions ($0.09 < \text{Cohen's}~d < 0.16$).

@@

As expected, Black participants ($M = 6.18, \textit{SD} = 1.31$) reported being more outraged about recent incidents of police violence than White participants ($M = 5.56, \textit{SD} = 1.61$; Cohen's $d = 0.42$). Black participants strongly identified with their ingroup ($M = 6.46, \textit{SD} = 1.09$) and, across experimental conditions, identified more with Black protesters ($M = 4.66, \textit{SD} = 1.99$) than with White protesters ($M = 4.03, \textit{SD} = 2.16$; $d = 0.30$). In contrast, White participants identified less strongly with their ingroup ($M = 4.85, \textit{SD} = 1.53$; $d = 1.21$) and did not identify more with White protesters ($M = 3.58, \textit{SD} = 2.01$) than with Black protesters ($M = 3.81, \textit{SD} = 1.77$; $d = -0.12$). Both White ($d = 0.67$) and Black ($d = 0.53$) participants tended to identify more with protesters protesting *for* defunding the police. On average, participants tended to describe their political orientation as moderately liberal ($M = 2.92, \textit{SD} = 1.73$) but, as expected, were divided in their support for defunding the police ($M = 3.37, \textit{SD} = 1.42$).

@@

Model 1 (M1) estimated varying intercepts for the eight conditions to test whether, in line with Hypotheses 1a, 1b, or 2a, participants' responses depended on their own group membership, the protesters' group membership, or the protesters' cause. Contradicting Hypothesis 1a, participants did not consider protest actions performed by their ingroup to be more acceptable, on average, than the same actions performed by the relevant outgroup (`r with(m1, d_text[contrast == "h1a_contrast"])`). Contradicting Hypothesis 1b, participants did not consider protest actions for a cause that was nominally aligned with their ingroup's interests to be more acceptable, on average, than protest actions for a cause not aligned with their ingroup's interests (`r with(m1, d_text[contrast == "h1b_contrast"])`). Contradicting Hypothesis 2a, participants did not consider protest actions for a system-defending cause to be more acceptable, on average, than protest actions for a system-challenging cause (`r with(m1, d_text[contrast == "h2a_contrast"])`). Instead, we found that, on average, White participants considered all protest actions to be more acceptable than Black participants (`r with(m1, d_text[contrast == "nh1_contrast"])`) and both Black (`r with(m1, d_text[contrast == "nh3_contrast"])`) and White (`r with(m1, d_text[contrast == "nh2_contrast"])`) participants considered the same actions to be more acceptable when protests were for, rather than against, defunding the police. As in Experiment 1, we thus found that participants' responses depended on both the participants' and the protesters' group memberships---but not in the directions predicted by our hypotheses.

@@

Model 2 (M2) extended Model 1 by estimating participants' responses as a function of their *z*-standardized endorsement of system-justifying beliefs. As preregistered, we modeled this relationship with two fixed effects, one estimating the effect of system-justifying beliefs on judgements about system-defending protest actions and one estimating their effect on judgements about system-challenging protest actions, and one varying effect estimating its variance across conditions. Supporting Hypothesis 2b, participants who *rejected* system-justifying beliefs were *more* likely to consider system-challenging protest actions (for defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "for"])`). We did not, however, find evidence for the ideological symmetry implied by Hypothesis 2b since participants who *endorsed* system-justifying beliefs were *not* more likely to consider system-defending protest actions (against defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "against"])`).

Figure \ref{fig:f5} shows the estimated pattern of condition-wise differences underlying those fixed effects. Participants who endorsed system-justifying beliefs tended to consider system-challenging and system-defending protest actions to be equally unacceptable. In contrast, participants who rejected system-justifying beliefs evinced ideology-based double standards: In line with Hypothesis 2b, they considered the same protest actions to be more acceptable when the protesters challenged the system or, to a lesser extent, when the protesters were from the disadvantaged group.

@@

Model 3 (M3) extended Model 1 by estimating participants' responses as a function of their self-reported support for the cause of the protest. We recoded participants' responses to create a predictor variable that encoded support for defunding the police when protesters supported defunding the police and opposition to defunding the police when protesters opposed defunding the police. As preregistered, we modeled this relationship as a monotonic effect [@burkner_modelling_2020] that estimated the average change in the outcome variable across predictor categories as well as how much of this change occurred between each of the four pairs of adjacent predictor categories.^[Our model assigned a Dirichlet prior, $\alpha = {1, 1, 1, 1}$, to the proportions of the overall change that was expected to occur between each of the four pairs of predictor categories.] Contradicting Hypothesis 3, participants who supported the protesters' cause did not, on average, consider the same protest actions to be more acceptable than participants who opposed the protesters' cause (`r with(m3_h3, b_text)`). Ergo, our results did not support the alternative hypothesis that ideology-based double standards can be reduced to support for, or opposition to, the cause of a protest.

@@

@@

We reran Model 2 with political orientation, instead of system justification, as an alternative predictor variable. Our results mirrored results from the preregistered analyses: More liberal participants were more likely to consider protest actions for defunding the police to be acceptable (`r with(m4_exp, b_text[Cause == "for"])`) but more conservative participants were not more likely to consider protest actions against defunding the police to be acceptable (`r with(m4_exp, b_text[Cause == "against"])`). As Figure 5 shows, conservative participants tended to consider all protest actions to be equally unacceptable while liberal participants considered the same protest actions to be more acceptable when protesters rallied around a progressive cause. Our non-preregistered analyses thus replicated the ideology-based double standards from the preregistered analyses with a different operationalization of political ideology.

@@

\ldots

@@

>>>>

---
title: "Double Standards in Judging Collective Action"
header: "DOUBLE STANDARDS"
keywords: protest, collective action, double standards, social identity theory, system justification theory, Black Lives Matter
abstract: "Collective action is a powerful force driving social change but often sparks contention about what actions are acceptable means to effect social change. In four studies (total *N* = 2,246), we investigated double standards in judging collective action---that is, whether observers will judge the same protest actions as more acceptable if the protesters are ingroup members and their cause aligns with the ingroup's interests (*identity-based double standards*) or if the protesters' cause aligns with the observers' ideological positions (*ideology-based double standards*). In two studies, we developed an instrument of 25 controversial protest actions, based in item response theory, to measure where people draw the line between acceptable and unacceptable forms of collective action. In two preregistered experiments, we used this instrument to test for double standards in judging collective action for workers' rights in the United Kingdom and for or against defunding the police in the United States. We found evidence for ideology-based, but not for identity-based, double standards: Participants judged the same protest actions to be more acceptable when the protesters' cause aligned with their own (system-challenging) ideological position---but showed no ingroup bias when judging collective action by ingroup and outgroup members. Our findings have theoretical and practical implications for understanding the often divided response to prominent social movements."
bibliography: references.bib
csl: apa.csl
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: yes
    template: preprint-template.tex

@@

Protest movements often spark contention about what actions are acceptable means of protest. Reactions to Black American athletes kneeling during the national anthem to protest racist police violence are a case in point. Kneeling during the national anthem is not violent, disruptive, or illegal. And yet, only 29\% of White Americans (66\% of Black Americans) and 11\% of Republicans (59\% of Democrats) considered it appropriate for Black athletes to protest in this way [@yougov_nfl_2017]. Indeed, Black athletes who had participated in the protests faced a backlash in public opinion [@lacina_nfl_2020]. 
<!-- Likewise, 42\% of Americans thought that, in the 2020 Black Lives Matter protests, most protesters were trying to incite violence or destroy property, even though very few protests involved any violence or property destruction [@acled_year_2021]. -->
This and other examples suggest that *who* the protesters are and *what* they are protesting influences how acceptable observers judge their actions to be.

@@

The second purpose of our research is to test for double standards in judging collective action---that is, whether people judge the same protest actions as more acceptable if the protesters' are ingroup members or the protesters' cause aligns with the ingroup's interests (*Hypothesis 1a*, *1b*) or if the protesters' cause aligns with their own system-justifying or system-challenging motives (*Hypotheses 2a*, *2b*). In Experiment 1, we use the instrument developed in Studies 1 and 2 to test for double standards in how acceptable participants consider collective action for workers' rights in the United Kingdom. By varying both the participants' and the protesters' group memberships (working/middle class), Experiment 1 compares the relative evidence for identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*). Extending Experiment 1, Experiment 2 considers both system-challenging and system-defending collective action in the context of the 2020 Black Lives Matter protests in the United States. By varying the protesters' cause (for/against defunding the police) as well as the participants' and the protesters' group memberships (Black/White), Experiment 2 compares the relative evidence identity-based (*Hypothesis 1a*, *1b*) and ideology-based double standards (*Hypothesis 2a*, *2b*). Together, the preregistered experiments provide a comprehensive test of the hypothesized double standards in judging collective action.

@@

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that, in line with *Hypothesis 1a* and *1b*, participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights than if they were outgroup members protesting for the outgroup's rights. By comparing reactions to working-class and middle-class protesters, we tested how observers judge collective action by lower-status (i.e. working-class) and higher-status (middle-class) group members. Assuming that collective action by lower-status group members is seen as more threatening to the unequal status quo than collective action by higher-status group members and that people are, to same extent, motivated to justify the system, we tested the preregistered hypothesis that, in line with *Hypothesis 2a*, participants would judge the same protest actions as more acceptable if the protesters were from a higher-status group than if they were from a lower-status group. In non-preregistered analyses, we further explored whether, in line with *Hypothesis 2b*, participants would judge the same protest actions as more or less acceptable depending on their own ideological positions. In this way, Experiment 1 provided a first test of the hypothesized identity- and ideology-based double standards in judging collective action.

@@

We preregistered the sample size as well as all hypotheses, inclusion/exclusion criteria, measures, and manipulations ([https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b](https://osf.io/24wrx/?view_only=8560eb60286149498f8da48bc9f9d99b)). We made all materials, data, and analysis scripts available online ([https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)). We followed sample size recommendations for item response theory models [@demars_irt_2010, p. 36], planning to recruit 500 participants.

@@

We recruited 500 participants from the remaining 687 participants, 250 from the lower-status group and 250 from the higher-status group. Participants were randomly assigned to read a vignette about a government bill affecting either people in working-class jobs (lower-status protesters) or people in professional jobs (higher-status protesters). Participants in both conditions were instructed to carefully read the vignette and to try to imagine what it would be like if this situation was real. Participants in the lower-status protesters condition read the following introduction:

@@

Figure \ref{fig:f3} (A) shows estimates for each combination of the protesters' and the participants' group membership. Overall, we found that participants' responses depended on both the protesters' and the participants' group status---but not in the directions predicted by our hypotheses. Contradicting our first hypothesis, participants did not consider protest actions performed by their ingroup to be more acceptable, on average, than the same actions performed by the relevant outgroup (Cohen's $d = 0.00, [-0.20, 0.20]$). Instead, we found that participants from higher-status backgrounds considered protest actions performed by both lower-status ($d = 0.34, [0.07, 0.59$]) and higher-status ($d = 0.33, [0.03, 0.63]$) protesters to be, on average, more acceptable than participants from lower-status backgrounds. Contradicting our second hypothesis, participants considered protest actions performed by higher-status protesters to be less acceptable, on average, than the same actions performed by lower-status protesters ($d = -0.20, [-0.40, -0.00]$).

@@

Experiment 1 provided a first test of the hypothesized identity- and ideology-based double standards in judging collective action. That is, we tested whether, in line with *Hypothesis 1a* and *1b*, participants would judge the same protest actions as more acceptable if the protesters were ingroup members protesting for the ingroup's rights or whether, in line with *Hypothesis 2a*, participants would judge the same protest actions as more acceptable if the protesters were from a higher-status group. We found that participants' responses depended on both the participants' and the protesters' group memberships---but not in the hypothesized directions. Instead, we found that both lower-status and higher-status participants tended to find the same protest actions more acceptable when lower-status protesters protested a bill threatening their rights. In non-preregistered analyses, we further found that people who were right-wing and endorsed system-justifying beliefs tended to find all protest actions to be less acceptable.

Experiment 1 did not, however, provide a conclusive test of our hypotheses. First, by focusing on fictitious government bills, we might have chosen a scenario too far removed from current affairs to evoke strong ingroup bias in participants' judgements of protest actions. That said, participants' reactions to the experimental manipulation suggested that it evoked outrage and was understood in ingroup--outgroup terms. Second, by focusing on collective action to defend workers' rights, we arguably studied reactions to collective action for a progressive cause. To test for ideology-based double standards, we relied on the assumption that collective action by lower-status group members would be seen as more challenging to the prevailing system than collective action by higher-status group members. This assumption, however, was contradicted by our observation that participants who endorsed system-justifying beliefs tended to find actions by lower-status and higher-status protesters to be less acceptable. Experiment 2 addressed those limitations by focusing on a scenario we expected to evoke stronger responses from participants and by examining reactions to both system-challenging and system-defending collective action.

@@

In Experiment 2, we investigated potential double standards in judging collective action for or against defunding the police in the United States. That is, we tested whether Black and White Americans applied different standards when judging collective action taken by either Black or White protesters to protest either for or against police divestment as a possible solution to racist police violence. We conducted this study in January 2021, a time when, after a year of unprecedented protests for racial justice, most Americans could be expected to be aware of, and have formed an opinion on, the Black Lives Matter movement. We focused on defunding the police as this position remained controversial among both Black and White Americans even as they broadly agreed on the need for police reform.^[In December 2020, only 34% of Black Americans and 18% of White Americans supported defunding the police while majorities of Black and White Americans supported police reforms such as eliminating qualified immunity (50% and 57%) or banning chokeholds [79% and 61%, @yougov_police_2020].]

By varying the participants' and the protesters' group memberships, we tested the preregistered hypothesis that participants would judge the same protest actions as more acceptable if the protesters were ingroup members (*Hypothesis 1a*). By assuming that defunding the police aligns more closely with the interests of Black than White Americans, we tested the preregistered hypothesis that participants would judge the same protest actions as more acceptable if the cause of the protest aligned with their ingroup's interests (*Hypothesis 1b*). By varying whether the protesters protested for or against defunding the police, we tested the preregistered hypotheses that participants would judge the same protest actions as more acceptable if the protest supported, rather than challenged, the system (*Hypothesis 2a*) and if they endorsed system-justifying beliefs and the protest supported the system or if they rejected system-justifying beliefs and the protest challenged the system (*Hypothesis 2b*). In this way, Experiment 2 provided a complete test of the hypothesized identity- and ideology-based double standards in judging collective action. In addition, we tested the simpler, alternative hypothesis that participants, in general, would judge the same protest actions as more acceptable if they supported the protesters' cause (*Hypothesis 3*).

@@

We preregistered the sample size as well as all hypotheses, inclusion/exclusion criteria, measures, and manipulations ([https://osf.io/skxjt/?view_only=d8ce44a700884b5ab36b64ef08f833a1](https://osf.io/skxjt/?view_only=d8ce44a700884b5ab36b64ef08f833a1)). We made all materials, data, and analysis scripts available online ([https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62](https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62)). As reported in the preregistration, we ran simulations, using data from Experiment 1, to determine that a sample size of $N = 1,600$ ($n = 200$ per condition) was sufficient to detect even small differences between conditions ($0.09 < \text{Cohen's}~d < 0.16$).

@@

As expected, Black participants ($M = 6.18, \textit{SD} = 1.31$) reported being more outraged about recent incidents of police violence than White participants ($M = 5.56, \textit{SD} = 1.61$; Cohen's $d = 0.42$). Black participants strongly identified with their ingroup ($M = 6.46, \textit{SD} = 1.09$) and, across experimental conditions, identified more with Black protesters ($M = 4.66, \textit{SD} = 1.99$) than with White protesters ($M = 4.03, \textit{SD} = 2.16$; $d = 0.30$). In contrast, White participants identified less strongly with their ingroup ($M = 4.85, \textit{SD} = 1.53$; $d = 1.21$) and did not identify more with White protesters ($M = 3.58, \textit{SD} = 2.01$) than with Black protesters ($M = 3.81, \textit{SD} = 1.77$; $d = -0.12$). Both White ($d = 0.67$) and Black ($d = 0.53$) participants tended to identify more with protesters protesting *for* defunding the police. On average, participants tended to describe their political orientation as moderately liberal ($M = 2.92, \textit{SD} = 1.73$) but, as expected, were divided in their support for defunding the police ($M = 3.37, \textit{SD} = 1.42$). That is, 56% supported defunding the police, 30% opposed defunding the police, and 14% remained undecided.

@@

Model 1 (M1) estimated varying intercepts for the eight conditions to test whether, in line with Hypotheses 1a, 1b, or 2a, participants' responses depended on their own group membership, the protesters' group membership, or the protesters' cause. Contradicting *Hypothesis 1a*, participants did not consider protest actions performed by their ingroup to be more acceptable, on average, than the same actions performed by the relevant outgroup (`r with(m1, d_text[contrast == "h1a_contrast"])`). Contradicting *Hypothesis 1b*, participants did not consider protest actions for a cause that was nominally aligned with their ingroup's interests to be more acceptable, on average, than protest actions for a cause not aligned with their ingroup's interests (`r with(m1, d_text[contrast == "h1b_contrast"])`). Contradicting *Hypothesis 2a*, participants did not consider protest actions for a system-defending cause to be more acceptable, on average, than protest actions for a system-challenging cause (`r with(m1, d_text[contrast == "h2a_contrast"])`). Instead, we found that, on average, White participants considered all protest actions to be more acceptable than Black participants (`r with(m1, d_text[contrast == "nh1_contrast"])`) and both Black (`r with(m1, d_text[contrast == "nh3_contrast"])`) and White (`r with(m1, d_text[contrast == "nh2_contrast"])`) participants considered the same actions to be more acceptable when protests were for, rather than against, defunding the police. As in Experiment 1, we thus found that participants' responses depended on both the participants' and the protesters' group memberships---but not in the directions predicted by our hypotheses.

@@

Model 2 (M2) extended Model 1 by estimating participants' responses as a function of their *z*-standardized endorsement of system-justifying beliefs. As preregistered, we modeled this relationship with two fixed effects, one estimating the effect of system-justifying beliefs on judgements about system-defending protest actions and one estimating their effect on judgements about system-challenging protest actions, and one varying effect estimating its variance across conditions. Supporting *Hypothesis 2b*, participants who *rejected* system-justifying beliefs were *more* likely to consider system-challenging protest actions (for defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "for"])`). We did not, however, find evidence for the ideological symmetry implied by *Hypothesis 2b* since participants who *endorsed* system-justifying beliefs were *not* more likely to consider system-defending protest actions (against defunding the police) to be acceptable means of protest (`r with(m2_h2, b_text[Cause == "against"])`).

Figure \ref{fig:f5} shows the estimated pattern of condition-wise differences underlying those fixed effects. Participants who endorsed system-justifying beliefs tended to consider system-challenging and system-defending protest actions to be equally unacceptable. In contrast, participants who rejected system-justifying beliefs evinced ideology-based double standards: In line with *Hypothesis 2b*, they considered the same protest actions to be more acceptable when the protesters challenged the system or, to a lesser extent, when the protesters were from the disadvantaged group.

@@

Model 3 (M3) extended Model 1 by estimating participants' responses as a function of their self-reported support for the cause of the protest. We recoded participants' responses to create a predictor variable that encoded support for defunding the police when protesters supported defunding the police and opposition to defunding the police when protesters opposed defunding the police. As preregistered, we modeled this relationship as a monotonic effect [@burkner_modelling_2020] that estimated the average change in the outcome variable across predictor categories as well as how much of this change occurred between each of the four pairs of adjacent predictor categories.^[Our model assigned a Dirichlet prior, $\alpha = {1, 1, 1, 1}$, to the proportions of the overall change that was expected to occur between each of the four pairs of predictor categories.] Contradicting *Hypothesis 3*, participants who supported the protesters' cause did not, on average, consider the same protest actions to be more acceptable than participants who opposed the protesters' cause (`r with(m3_h3, b_text)`). Ergo, our results did not support the alternative hypothesis that ideology-based double standards can be reduced to support for, or opposition to, the cause of a protest.

@@

  # Import results
  m5 <- read_rds("../Experiment 2/results/m5_post.rds")

  # Define contrasts
  m5 <- m5 %>% 
    mutate(
      h1a_contrast = case_when(
        Participant == Protesters ~ +1L,
        Participant != Protesters ~ -1L,
      ),
      h1b_contrast = case_when(
        Participant == "Black" & Cause == "For" ~ +1L,
        Participant == "White" & Cause == "Against" ~ +1L,
        TRUE ~ -1L,
      )
    )

  # Calculate contrasts
  m5 <- m5 %>% 
    group_by(.draw) %>% 
    summarize(
      across(
        ends_with("_contrast"),
        ~mean(z * .)
      )
    ) %>% 
    pivot_longer(
      ends_with("_contrast"),
      names_to = "contrast",
      values_to = "d"
    ) %>% 
    group_by(contrast) %>% 
    median_qi(d)

  # Make text
  m5 <- m5 %>% 
    mutate(
      across(
        c(d, .lower, .upper), 
        f_num, digits = 2, retain.leading.zero = T
      ),
      d_text = glue(
        "$d = {d}, [{.lower}, {.upper}]$"
      )
    )

We explored whether group identification moderated how the participants' and the protesters' group memberships affected participants' responses. To that end, we extended Model 1 by estimating participants' responses as a function of their *z*-standardized identification with their racial ingroup. As in Model 2, we modeled this relationship with a fixed and a varying effect. We found, however, that even participants who strongly identified with their racial ingroup (+1*SD*) did not consider protest actions performed by their ingroup to be more acceptable than the same actions performed by the relevant outgroup (`r with(m5, d_text[contrast == "h1a_contrast"])`) and did not consider protest actions for a cause that was aligned with their ingroup's interests to be more acceptable  than protest actions for a cause not aligned with their ingroup's interests (`r with(m5, d_text[contrast == "h1b_contrast"])`). Our non-preregistered analyses thus suggested that group identification did not moderate group differences in judgements about collective action by ingroup and outgroup members.

@@

We also explored whether we would find ideology-based double standards when operationalizing ideology as political orientation instead of as system justification. To that end, we reran Model 2 with political orientation as the *z*-standardized predictor variable. Our results mirrored the preregistered analyses: More liberal participants were more likely to consider protest actions for defunding the police to be acceptable (`r with(m4_exp, b_text[Cause == "for"])`) but more conservative participants were not more likely to consider protest actions against defunding the police to be acceptable (`r with(m4_exp, b_text[Cause == "against"])`). As Figure 5 shows, conservative participants tended to consider all protest actions to be equally unacceptable while liberal participants considered the same protest actions to be more acceptable when protesters rallied around a progressive cause. Our non-preregistered analyses thus replicated the ideology-based double standards from the preregistered analyses with a different operationalization of ideology.

@@

Experiment 2 provided a complete test of the hypothesized identity- and ideology-based double standards in judging collective action. As in Experiment 1, we found that participants' responses depended on both the participants' and the protesters' group memberships---but not in the hypothesized directions. We found that, contrary to *Hypothesis 1a*, participants considered protest actions taken by ingroup and outgroup members to be equally acceptable and that, contrary to *Hypothesis 1b*, participants did not consider the same protest actions to be more acceptable when the protesters' cause aligned with their ingroup's interests. Like Experiment 1, Experiment 2 thus found no evidence for identity-based double standards---even though it focused on an issue of direct and current relevance to the participants.

Expanding Experiment 1, Experiment 2 considered reactions to both system-challenging and system-defending collective action and, in so doing, provided a stronger test of the hypothesized ideology-based double standards. Contrary to *Hypothesis 2a* and the underlying assumption of a universal system-justifying motive, we found that participants considered the same protest actions to be *more* acceptable when judging system-challenging collective action (for defunding the police) than when judging system-defending collective action. This finding might reflect our liberal-leaning sample. In line with *Hypothesis 2b*, we found that participants who rejected system-justifying beliefs considered the same protest action more acceptable when judging system-challenging collective action (for defunding the police). We did not, however, find evidence for ideological symmetry in this relationship as participants who endorsed system-justifying beliefs considered system-challenging and system-defending collective action to be equally unacceptable. In addition, we did not find evidence for the alternative, simpler *Hypothesis 3* that that participants would judge the same protest actions to be acceptable when they supported the protesters' cause. In this way, Experiment 2 provided evidence for ideology-based, but not for identity-based, double standards in judging collective action.

@@

Our research was motivated by real-world examples that seemed to suggest that where people draw the line between acceptable and unacceptable protest actions depends on *who* the protesters are and *what* they are protesting---in other words, that there are double standards in judging collective action. Our findings, however, showed that it is not so simple. Contradicting our hypothesis of identity-based double standards, participants in two preregistered experiments did not show ingroup bias when judging controversial protest actions taken by ingroup and outgroup members to advance their group's interests. Instead, supporting our hypothesis of ideology-based double standards, participants judged the same controversial protest actions to be more acceptable when the protesters' cause aligned with their own (system-challenging) ideological position. 
<!-- This suggests that where observers draw the line between acceptable and unacceptable forms of collective action depends not so much on the protesters' and the observers' identities, but on the protesters' cause and how it aligns with the observers' political and ideological orientation.  -->
In the remainder of this General Discussion, we discuss theoretical and practical implications for understanding the often divided response to collective action and consider some limiting conditions on our findings.

## Implications

<!-- Our findings raised the question why we did not find double standards in judging collective action, even though there are well-documents double standards in other domains. -->

<!-- 1. What does it mean that we did not find identity-based double standards? -->
<!-- We do not find ingroup bias, at least not for broad social categories such as race and class, in contrast to evidence and predictions from social identity theory. Protest actions are not the same as moral transgressions, perhaps. We might have reached a different conclusion when adopting the framework of opinion-based group memberships and understanding political ideology in partisan identity terms. -->
<!-- 2. What does it mean that we did not find symmetrical ideology-based double standards? -->
<!-- It could just be that conservatives don't like protest. But consider Jan. 6 --- maybe it's about the cause of the protest? We know about the distinct moral values of conservatives and liberals (see MFT, see MFT and political orientation) and how they shape attitudes toward policy issues (see MFT and culture war issues). This aligns with Verkuyten and Richardson's finding of morality and tolerance judgements. -->

Our results showed that, for many controversial protest actions, there is no consensus on what is and is not an acceptable means to advance a cause. Instead, this judgement seems to depend on what the cause of a protest is and how it aligns with one's beliefs and values. In this way, our results confirm our contention that the distinction between normative and non-normative forms of collective action should be situated in the eye of the beholder. By developing an instrument for measuring where people draw the line between acceptable and unacceptable forms of collective action, we provided a paradigm that we hope will stimulate research on when and why people differ in making this distinction.
